### 2021-07-05 rad-icfp-2021/intro.tex
minor
Intro until section 1.4 + US spelling

% % (Source-to-source) Automatic Differentiation (AD) is a technique for transforming code that implements 
% % a function $f$ into code that computes $f$'s derivative, essentially by using the chain rule for derivatives. 
% % Due to its efficiency and numerical stability, AD is the technique of choice whenever derivatives need to be computed 
% % of functions that are implemented as programs, particularly in high dimentional settings. 
% % Optimisations and Monte Carlo integration algorithms, such as Gradient Descent and Hamiltonian Monte-Carlo methods, 
% % rely crucially on the calculation of derivatives. These algorithms are used in virtually everywhere in machine learning 
% % and computational statistics, and the calculation of derivatives is usually the computational bottle-neck. 
% % AD, roughly speaking, comes in two modes; forward-mode and reverse-mode. 
% % When differentiating a function $\RR^n\to\RR^m$, forward-mode tends to be more efficient if $m \gg n$, while reverse-mode 
% % is generally more performant if $n \gg m$. As most applications reduce to optimization of Monte-Carlo 
% % integration of an objective function $\RR^n\to\RR$ with $n$ very large ($10^4-10^7$), reverse-mode AD is 
% % in many ways the more interesting algorithm.

@@

In 2008, Pearlmutter and Siskind~\cite{pearlmutter2008reverse} introduced Automatic Differentiation (AD) in a functional setting, where they introduced reverse-mode. 
Reverse-mode computes in a single fast pass the whole gradient of a function $\RR^n\to\RR$.
By comparison, the easier forward-mode transformation would need $n$ passes. 
These functions $\RR^n\to\RR$ appear a lot in practice: for instance, as loss functions in machine learning.
Reverse-mode is thus important, as it computes the whole gradient of such functions very quickly. 
However, in order to be as efficient as the usual imperative version of reverse-mode, the transformation in~\cite{pearlmutter2008reverse} resorted to references.

The lack of purity of reverse-mode is known to make it significantly harder to optimize and parallelize. 
None of the current implementations of reverse-mode in a functional setting~\cite{lantern_icfp,pearlmutter2008reverse,baydin2016diffsharp} are pure, and often complicated heuristics with no guarantees are used, e.g.~\cite{xla}.

This means that for every important non trivial operation, a hand-crafted reverse-derivative must be given, to optimize for efficiency.
In other words, abstracting away from imperative code is still a hurdle that functional implementations need to overcome.

% For example, \MH{TODO: show how $\prod_i v_i$ for a vector $v$ is done in some functional AD systems}.
% For real world AD systems, modularity is key. They need to be expressive but have a lot of optimised derivatives for operators that are used a lot.
% This is easily seen on numpy \MH{TODO: explain a bit better why torch is good by using optimised grad for numpy.}
% An efficient purely functional reverse-mode transformation would be more easily explainable, optimisable, and potentially reusable.
An efficient purely functional reverse-mode transformation would be more easily explainable, optimizable, and reusable.

TODO: rewrite for more theory people. some ideas: for canonical than Plotkin's thing because normal semantics, complexity proof compared to all other recent papers, 
all papers that have implementations, their interface in the paper is too high level and misses a lot of things (e.g. in \cite{vytiniotis2019differentiable,lantern_icfp,sherman2021}) compared to implementation, hinders progress. 
For instance, in \cite{lantern_icfp} it's a lot of manually optimized and hand-derived primitives to achieve good performance.

@@

This raises several issues that are dealt with throughout the paper. First, there is a tradeoff to reach between a general expressive language and a domain specific one. The latter usually has more static information and a specific representation that lends itself to better optimizations.
Then, a lot of optimizations performed on AD implementations consist of a lot of hand-crafted derivatives for a lot of useful operations like matrix-matrix multiplication, dot-product, etc.
They don't seem to come with more theoretical justification, are error prone, and can hardly fit more general optimizations.
\MH{TODO: maybe the example and ref that ML systems spend 90percent of the time outside of numpy.}
This makes these systems harder to prove correct. The problem is thus to stay provably correct and pure, and yet efficient. What's more is we would like something easily provably efficient. 
A real implementation on top of our work would of course optimize further, potentially using hand-crafted operations as well, but it would be based on more solid grounds.

TODO: same as above, maybe rewrite for more theory audience.

@@

We propose a source-code transformation on simple purely functional language for a purely functional reverse-mode differentiation.
Our transformation is decomposed in a sort of compilation scheme through a novel intermediate representation, unary form (UNF).
To emphasize some key ideas and for presentation purposes, we present our work with a simple yet expressive language in Section~\ref{sec:simplediff}, 
and show how to extend our work to a richer language in Section~\ref{sec:generalisation}.

@@

Our contribution makes these claims precise. In detail we introduce
\begin{itemize}
    \item an expressive source language with array operations, yet restrictive enough for efficient differentiation (\S\ref{sub:sourcelang})
    \item a method for extending our work and modular method for adding and prove correct new primitives as needed in practice (\S\ref{sec:generalisation})
    \item a new intermediate representation (I.R), UNF, essentially turning every operation to a unary one which inputs and outputs a tuple (\S\ref{sec:unf})
    \item a transformation to and from the I.R. from the main language (\S\ref{sub:transformations to and from UNF})
    \item a simple reverse-mode transformation on the I.R. (\S\ref{sub:Simple reverse mode transformation})
    \item the direct reverse-mode transformation obtained from the source to the target language (\S\ref{sub:Macro for pure reverse mode transformation})
    \item denotational semantics of our languages and transformations using multicategories and concategories (\S\ref{sec:correctness})
    \item correctness of the transformations (\S\ref{sec:correctness})
    \item complexity guarantees (\S\ref{sec:complexity})
    \item how efficiency is recovered via standard optimization techniques (\S\ref{sub:Optimisations})
    \item several additions to the source language (\S\ref{sec:generalisation}) \MH{should emphasize that it should be easier for us to add primitives compared to several recent work. give recipe.}
    \item a detailed comparison with other work and some limitations to our work (\S\ref{sec:conclusion})
\end{itemize}

@@

In Section~\ref{sec:background} we recall rudiments of automatic differentiation, forward and reverse-mode differentiation.
Next, in Section~\ref{sec:simplediff} we introduce the source and target languages for differentiation, and our reverse mode transformation. 
In Section~\ref{sec:unf} we introduce a new intermediate representation UNF, and a simple reverse-mode macro on this representation.
We introduce transformations from the source language to UNF and from UNF to the target language.
Next, in Section~\ref{sec:correctness} we give denotational semantics to our languages and show that our decomposition respects the diagram above and that our reverse mode transformation is correct.
In Section~\ref{sec:complexity}, we show that our transformation verifies the cheap gradient principle and extra general optimisations.
In Section~\ref{sec:generalisation} we show how to adapt our work for a richer source language. 
Finally, related work, limitations to our approach and future work are presented in Section~\ref{sec:conclusion}.

>>>>

% % (Source-to-source) Automatic Differentiation (AD) is a technique for transforming code that implements 
% % a function $f$ into code that computes $f$'s derivative, essentially by using the chain rule for derivatives. 
% % Due to its efficiency and numerical stability, AD is the technique of choice whenever derivatives need to be computed 
% % of functions that are implemented as programs, particularly in high dimensional settings. 
% % Optimizations and Monte Carlo integration algorithms, such as Gradient Descent and Hamiltonian Monte-Carlo methods, 
% % rely crucially on the calculation of derivatives. These algorithms are used in virtually everywhere in machine learning 
% % and computational statistics, and the calculation of derivatives is usually the computational bottle-neck. 
% % AD, roughly speaking, comes in two modes; forward-mode and reverse-mode. 
% % When differentiating a function $\RR^n\to\RR^m$, forward-mode tends to be more efficient if $m \gg n$, while reverse-mode 
% % is generally more performant if $n \gg m$. As most applications reduce to optimization of Monte-Carlo 
% % integration of an objective function $\RR^n\to\RR$ with $n$ very large ($10^4-10^7$), reverse-mode AD is 
% % in many ways the more interesting algorithm.

@@

In 2008, Pearlmutter and Siskind~\cite{pearlmutter2008reverse} introduced Automatic Differentiation (AD) in a functional setting, where they defined reverse-mode. 
Reverse-mode computes in a single fast pass the whole gradient of a function $\RR^n\to\RR$.
By comparison, the easier forward-mode transformation would require $n$ passes. 
These functions $\RR^n\to\RR$ appear a lot in practice: for instance, as loss functions in machine learning.
Reverse-mode is thus important, as it computes the whole gradient of such functions very quickly. 
However, in order to reach the efficiency of the usual imperative version of reverse-mode, the transformation in~\cite{pearlmutter2008reverse} resorts to references.

The lack of purity in reverse-mode is known to make it significantly harder to optimize and parallelize. 
None of the current implementations of reverse-mode in a functional setting~\cite{lantern_icfp,pearlmutter2008reverse,baydin2016diffsharp} are pure, and often complicated heuristics with no guarantees are used, e.g.~\cite{xla}.

As a result, to optimize for efficiency, a hand-crafted reverse-derivative must be given, for every important non trivial operation.
In other words, abstracting away from imperative code is still a hurdle that functional implementations need to overcome.

An efficient purely functional reverse-mode transformation would be more easily explainable, optimizable, and reusable.

\MH{TODO: rewrite for more theory people. some ideas: for canonical than Plotkin's thing because normal semantics, complexity proof compared to all other recent papers, 
all papers that have implementations, their interface in the paper is too high level and misses a lot of things (e.g. in \cite{vytiniotis2019differentiable,lantern_icfp,sherman2021}) compared to implementation, hinders progress. 
For instance, in \cite{lantern_icfp} it's a lot of manually optimized and hand-derived primitives to achieve good performance.}

@@

This raises several issues that we deal with throughout the paper. First, there is a tradeoff to reach between a general expressive language and a domain specific one. The latter usually has more static information and a specific representation that lends itself to better optimizations.
Then, many optimizations performed on AD implementations consist in hand-crafted derivatives for useful operations like matrix-matrix multiplication, dot-product, etc.
They don't seem to arise from theoretical justifications, are error prone, and can hardly fit with more general optimizations.
\MH{TODO: maybe the example and ref that ML systems spend 90percent of the time outside of numpy.}
This makes these systems harder to prove correct. The problem is thus to ensure provable correctness and pureness, while not compromising on efficiency. In addition, we would like something easily provably efficient. 
A real-world implementation based on our work would of course optimize further, potentially using hand-crafted operations as well, but it would be based on solid grounds.

\MH{TODO: same as above, maybe rewrite for more theory audience.}

@@

We propose a source-code transformation on a simple purely functional language for purely functional reverse-mode differentiation.
Our transformation is comprised of compilation scheme through a novel intermediate representation: unary form (UNF).
To emphasize the key ideas and for expository purposes, we present our work with a simple yet expressive language in Section~\ref{sec:simplediff}, 
and show how to extend our work to a richer language in Section~\ref{sec:generalisation}.

@@

Our contribution makes these claims precise. In more details, we introduce
\begin{itemize}
    \item an expressive source language with array operations, yet restrictive enough for efficient differentiation (\S\ref{sub:sourcelang})
    \item a method for extending our work and modular approaches to add and prove correct new primitives as needed in practice (\S\ref{sec:generalisation})
    \item a new intermediate representation (IR), UNF, essentially turning every operation into a unary one which inputs and outputs a tuple (\S\ref{sec:unf})
    \item a transformation between the IR and the main language (\S\ref{sub:transformations to and from UNF})
    \item a simple reverse-mode transformation on the IR (\S\ref{sub:Simple reverse mode transformation})
    \item the direct reverse-mode transformation obtained from the source to the target language (\S\ref{sub:Macro for pure reverse mode transformation})
    \item denotational semantics of our languages and transformations using multicategories and concategories (\S\ref{sec:correctness})
    \item correctness of the transformations (\S\ref{sec:correctness})
    \item complexity guarantees (\S\ref{sec:complexity})
    \item how efficiency is recovered via standard optimization techniques (\S\ref{sub:Optimizations})
    \item several additions to the source language (\S\ref{sec:generalisation}) \MH{should emphasize that it should be easier for us to add primitives compared to several recent work. give recipe.}
    \item a detailed comparison with other work and some limitations to our work (\S\ref{sec:conclusion})
\end{itemize}

@@

In Section~\ref{sec:background} we recall rudiments of automatic differentiation, forward and reverse-mode differentiation.
Next, in Section~\ref{sec:simplediff} we introduce the source and target languages for differentiation, and our reverse mode transformation. 
In Section~\ref{sec:unf} we introduce a new intermediate representation UNF, and a simple reverse-mode macro on this representation.
We introduce transformations from the source language to UNF and from UNF to the target language.
Next, in Section~\ref{sec:correctness} we give denotational semantics to our languages and show that our decomposition respects the diagram above and that our reverse mode transformation is correct.
In Section~\ref{sec:complexity}, we show that our transformation verifies the cheap gradient principle and extra general optimizations.
In Section~\ref{sec:generalisation} we show how to adapt our work for a richer source language. 
Finally, related work, limitations to our approach and future work are presented in Section~\ref{sec:conclusion}.

### 2021-07-05 rad-icfp-2021/relwork.tex
Intro until section 1.4 + US spelling

We have introduced a transformation on programs to compute provably efficient (\ref{sec:complexity}) 
gradients via reverse derivatives in a purely functional way (\ref{sub:Macro for pure reverse mode transformation})
on a simple yet expressive language with functions on arrays (\ref{sub:sourcelang}), 
combined with standard functional optimisations (\ref{fig:optim}).  
We introduced a novel intermediate representation, unary form (\ref{sec:unf}) 
to decompose our transformation into simpler transformations.
We gave denotational semantics to our languages and our transformations (\ref{sec:correctness}) 
and showed correctness of the reverse-mode transformation (\ref{sub:Correctness theorem}).
We showed (\ref{sec:generalisation}) how to lift some of the restrictions that
we introduced on arrays and how to extend our approach to a richer language with more primitive operations and conditionals.

@@

\noindent \textbf{More dynamic language}
JAX \cite{bradbury2020jax,frostig2018compiling} is a modern framework that provides state-of-the-art automatic differentiation. 
One feature that makes it popular is that it uses JIT compilation, 
which seems to be the best avenue to help best ML practionners as it allows them 
to quickly experiment but also benefit from the full power of static optimisations.
We are interested in seeing how our work could sit in a JIT compiler.

\noindent \textbf{More optimisations.}
As pointed in \ref{sub:Adding general arrays} it would be interesting to have a more principled representation 
of tensors which could lead to more optimisations. 
There is always a tension between dynamicity and the possibility for static optimisations 
and we believe our language offers a good tradeoff, but it will be interesting to tailor it 
further for specific applications.

@@

\noindent \textbf{Parallelism and GPU}
We tried to locally preserve parallelism. Having a pure transformation should help with found parallelism. 
It would be interesting to implement on transformations and see how they perform on GPU.
It is well-known that the addition of references provoked by the standard reverse-mode transformation
restricts a lot of optimizations and it is not hard to find examples in C where the compiler does not detect simple optimisations 
because of the references introduced by reverse-mode.

>>>>

We have introduced a transformation on programs to compute provably efficient (\ref{sec:complexity}) 
gradients via reverse derivatives in a purely functional way (\ref{sub:Macro for pure reverse mode transformation})
on a simple yet expressive language with functions on arrays (\ref{sub:sourcelang}), 
combined with standard functional optimizations (\ref{fig:optim}).  
We introduced a novel intermediate representation, unary form (\ref{sec:unf}) 
to decompose our transformation into simpler transformations.
We gave denotational semantics to our languages and our transformations (\ref{sec:correctness}) 
and showed correctness of the reverse-mode transformation (\ref{sub:Correctness theorem}).
We showed (\ref{sec:generalisation}) how to lift some of the restrictions that
we introduced on arrays and how to extend our approach to a richer language with more primitive operations and conditionals.

@@

\noindent \textbf{More dynamic language}
JAX \cite{bradbury2020jax,frostig2018compiling} is a modern framework that provides state-of-the-art automatic differentiation. 
One feature that makes it popular is that it uses JIT compilation, 
which seems to be the best avenue to help best ML practitioners as it allows them 
to quickly experiment but also benefit from the full power of static optimizations.
We are interested in seeing how our work could sit in a JIT compiler.

\noindent \textbf{More optimizations.}
As pointed in \ref{sub:Adding general arrays} it would be interesting to have a more principled representation 
of tensors which could lead to more optimizations. 
There is always a tension between dynamicity and the possibility for static optimizations 
and we believe our language offers a good tradeoff, but it will be interesting to tailor it 
further for specific applications.

@@

\noindent \textbf{Parallelism and GPU}
We tried to locally preserve parallelism. Having a pure transformation should help with found parallelism. 
It would be interesting to implement on transformations and see how they perform on GPU.
It is well-known that the addition of references provoked by the standard reverse-mode transformation
restricts a lot of optimizations and it is not hard to find examples in C where the compiler does not detect simple optimizations 
because of the references introduced by reverse-mode.
